---
title: "Daily Exercise 8"
subtitle: "Practice Making Joins and Pivots"
author:
- name: Josh Puyear
  email: "jpuyear@colostate.edu"
format: html
execute: 
  echo: true
---

# Make a faceted plot of the cumulative cases & deaths by USA region. 
Your x axis should be the date and the y axis value/count. To do this you will need to join and pivot the COVID-19 data. We can break this task down:

# 1. Read in the COVID-19 Data
```{r, echo = TRUE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)


url = 'https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-recent.csv'
covid = read_csv(url)
head(covid, 5)
```

# 2. Create a new data.frame using the available state.abb, state.name, state.region objects in  base R.
Be intentional about creating a primary key to match to the COVID data!

```{r, echo = TRUE}
df <- data.frame(region = state.region,
                 abbr = state.abb,
                 state = state.name)
head(df)
```

# 3. Join your new data.frame to the raw COVID data. 
Think about right, inner, left, or full join…

```{r, echo = TRUE}
covid_region <- inner_join(df, covid, by = "state") |>
  group_by(region, date) |>  
  summarize(cases = sum(cases),  

            deaths = sum(deaths), 
            .groups = "drop") |>  
# Remove grouping to prevent issues
  pivot_longer(cols = c(cases, deaths),
               names_to = "type",
               values_to = "count") |>
  ggplot(aes(x = date, y = count, group = type, color = type)) +  # Ensure grouping
  geom_line() +                 
  facet_grid(type ~ region) +
  theme_bw()
```


# 4. Save the image to your img directory with a good file name and extension!

```{r, echo = TRUE}
ggsave("covid_cases.png", plot = covid_region, width = 10, height = 6, dpi = 300, 
       path = "C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ess-330-daily-exercises/all-daily-exercises/imgs")
```