---
title: "Projects"
format: html
editor: visual
output-dir: docs
---

#This activity is to demonstrate training and testing data for models.

#Accessing Palmer Penguins Dataset

```{r, echo = TRUE}
#accessing Palmer Penguins

library(tidymodels)
library(palmerpenguins)


set.seed(3155)
(resample_split <- initial_split(penguins, prop = 0.7))

```

# Creating Training Data

```{r, echo = TRUE}
#create training data
penguins_train <- training(resample_split)
glimpse(penguins_train)

```

#Creating Testing Data

```{r, echo = TRUE}
#create testing data
penguins_test <- testing(resample_split)
glimpse(penguins_test)
```

# Assessing Data Proportions

```{r, echo = TRUE}
#Checking that data proportions match

(table(penguins$species) / nrow(penguins))

# Training Data
(table(penguins_train$species) / nrow(penguins_train))

# Testing Data
(table(penguins_test$species) / nrow(penguins_test))

#They don't really match
```

# Stratification makes the proportions even for each subset of the data

```{r, echo = TRUE}
#Making strata
set.seed(142)

penguins <- drop_na(penguins)
penguins_strata <- initial_split(penguins, strata = species, prop = .7)

```

#Extract training and testing datasets from stratified data

```{r, echo = TRUE}

# Extract the training and testing sets
train_strata <- training(penguins_strata)
test_strata  <- testing(penguins_strata)

```

#Making Sure Proportions are more aligned

```{r, echo = TRUE}
# Check the proportions
# Dataset
table(penguins$species) / nrow(penguins)

table(train_strata$species) / nrow(train_strata)

table(test_strata$species) / nrow(test_strata)

```

#Create a 10-fold cross validation based on the training data

```{r, echo = TRUE}
#Create a 10-fold cross validation based on the training data

#this is how many rows will be in each fold
nrow(train_strata) * 1/10

vfold_cv(train_strata, v = 10) |>
  glimpse()


```
