---
title: "Daily Exercise 21"
subtitle: "Time Series Practice"
author:
  name: Josh Puyear
  email: "jpuyear@colostate.edu"
editor: visual
project:
  output-dir: docs
format: html
execute: 
  echo: true
self-contained: true

---

# Downloading USGS data

```{r, echo = FALSE}
library(dataRetrieval)
library(dplyr)
library(tidyr)
library(lubridate)
library(plotly)
library(tidyverse)
library(zoo)
library(feasts)
library(tsibble)
library(ggpubr)
library(ggplot2)

# Example: Cache la Poudre River at Mouth (USGS site 06752260)
poudre_flow <- readNWISdv(siteNumber = "06752260",    # Download data from USGS for site 06752260
                          parameterCd = "00060",      # Parameter code 00060 = discharge in cfs)
                          startDate = "2013-01-01",   # Set the start date
                          endDate = "2023-12-31") |>  # Set the end date
  renameNWISColumns() |>                              # Rename columns to standard names (e.g., "Flow", "Date")
  mutate(Date = yearmonth(Date)) |>                   # Convert daily Date values into a year-month format (e.g., "2023 Jan")
  group_by(Date) |>                                   # Group the data by the new monthly Date
  summarise(Flow = mean(Flow))                       # Calculate the average daily flow for each month

```


# 1. Convert to tsibble
```{r, echo = TRUE}
poudre_flow <- as_tsibble(poudre_flow)

head(poudre_flow)
```

# 2. Plotting time series
```{r, echo = TRUE}
poudre_plot <- ggscatter(poudre_flow, x = "Date", y = "Flow")+
  geom_line(alpha = 0.5)

ggplotly(poudre_plot)

```

# 3. Subseries
```{r, echo = TRUE}
gg_subseries(poudre_flow) +
  labs(title = "Seasonal Poudre Flow Patterns",
       y = "Average Flow",
       x = "Date (Year)") +
  theme_minimal()


gg_season(poudre_flow)+
  labs(title = "Seasonal Poudre Flow Patterns",
       y = "Average Flow",
       x = "Month")

```
## Answer to Question 3
In this graphic, seasons are defined by the month. A subseries is the history of each month plotted through multiple years.

# Decomposing the Data
```{r, echo = TRUE}
poudre_decomp <- poudre_flow |>
  model(STL(Flow ~ season(window = "periodic"))) |>
  components()

glimpse(poudre_decomp)

poudre_trend <- ggscatter(poudre_decomp, y = "trend", x = "Date")

poudre_season <- ggscatter(poudre_decomp, y = "season_year", x = "Date")

```
## Answer 4.
What I see is a gradual decrease in the Poudre's flow like a bear market trend. This could be reflective of increasing pressures on the Poudre watershed in the last decade from development and drought. Plotting season_year, trends are removed through time to just reveal seasonal patterns. With this graph, the patterns are very symmetrical, suggesting that season has a predicatble influence on river flow and there are other factors at play outside of season.


